Local Search Approximation Algorithms for Partitional and Hierarchical Clustering

### 2015-05-10 tex/project.tex
just some typo from daves edit

There are two main branches of clustering: partitional (or flat) and hierarchical. In partitional clustering we want to partition the data into disjoint sets whereas in hierarchical clustering we are interested in finding a nested tree of partitions. Often these problems are solved using heuristic algorithms which have no guarantee on the solution produced but are typically fast and easy to implement. We will look at how local search approximation algorithms, which do have some approximation guarantee on the constructed solution, and are less computationally expensive than other approximation methods, compare in practice to some of the most widely used clustering algorithms both in the quality of the solution and in computation time.

@@

Arguably the most common partitional clustering method is $k$-means, in which we want to choose $k$ cluster centers to minimize the sum of squared distances from each point to its cluster center \cite{Jain}.  $k$-means is usually solved using Lloyd's algorithm (also referred to simply as the $k$-means algorithm), which iteratively greedily updates the centers of each cluster (assigning points to the closest center) until the solution becomes stable \cite{Jain}. Lloyd's algorithm is only guaranteed to reach a local optimum and can be very sensitive to outliers and the initial points chosen \cite{Kanungo}.

@@

Various approximation algorithms for he $k$-median problem have been proposed, with a best known approximation ratio of 2.592 \cite{Wu}. However, this algorithm uses linear programming and is computationally intensive compared to local search and greedy algorithms.  It is known that a local search algorithm that considers swaps of size $\leq p$ is a $(3 + \frac{2}{p})$-approximation algorithm \cite{Arya}. Note that for $p=1$, this equivalent to PAM.  As far as we know, this algorithm has not been implemented or tested on clustering data, so the relative trade-off of solution quality and computation time as $p$ increases is unknown.

@@

The most common method for hierarchical clustering is agglomerative clustering. Agglomerative clustering starts with each data point in its own cluster, and proceeds to recursively merge the two closest clusters to form the next clustering level in the hierarchy, repeating until the entire data set is a single cluster.  There are several choices on how to measure distance in order to choose which clusters to merge. Single linkage (also called nearest neighbor) measures the distance between two clusters to be the minimum pairwise distance between the data points in each cluster.  Complete linkage clustering instead defines the distance to be the largest pairwise distance. A method between these two extremes is group average linkage which looks at the average pairwise distance between clusters. This strikes a balance between the closeness and compactness of the two techniques above and gives more reasonable results in practice \cite{ESL}. Lastly, there is Ward's method which chooses the two clusters to merge so as to minimize the sum of each cluster's variance \cite{Ward}.

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. In particular, their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. The local search $(3+\frac{2}{p})$-approximation algorithm as the $k$-median solver performed well in experimental tests \cite{Nagarajan}. While the authors tested the variants of this algorithm against other $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

@@

\begin{figure*}[h]
\label{fig:flat_results}
\small
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Time & Time & Time & Time & Time & Time & Time & Time & Time \\ \hline
	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap  \\ \hline
	Soybean & 0.015 & 0.061 & 0.002 & 1.179 & 0.026 & 0.016 & 0.048 & 0.049 & 0.081 \\ 
	Iris & 0.013 & 0.023 & 0.003 & 9.788 & 0.125 & 0.102 & 0.212 & 0.229 & 0.414 \\ 
	Gauss\_5\_5\_5 & 0.025 & 0.056 & 0.032 & 75.262 & 1.159 & 1.506 & 2.188 & 2.382 & 3.71 \\ 
	Gauss\_5\_10\_5 & 0.054 & 0.095 & 0.109 & 134.137 & 6.834 & 5.407 & 9.243 & 9.428 & 15.96 \\ 
	%Gauss\_5\_50\_5 & 0.137 & 0.196 & 0.302 & 97.336 & 27.856 & 21.446 & 32.407 & 39.614 & 44.69 \\ 
	Gaussian avg & 0.073 & 0.128 & 0.182 & 144.455 & 13.289 & 13.021 & 18.749 & 18.696 & 24.592 \\ 
	pmed5 & 0.038 & 8.298 & 0.481 & 0.622 & 0.653 & 0.748 & 0.939 & \  & \  \\ 
	pmed20 & 1.795 & 174.951 & 52.208 & 44.979 & 68.554 & 68.231 & 79.288 & \  & \  \\ 
	% pmed40 & 3.527 & 849.852 & 251.259 & 264.867 & 250.412 & 391.738 & 350.235 & \  & \  \\ 
	pmedian avg & 0.807 & 203.308 & 40.386 & 40.347 & 48.264 & 56.76 & 56.19 & \  & \  \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI  \\ \hline
	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap   \\ \hline
	Soybean & 0.686 & 0.178 & 0.55 & 0.612 & 0.708 & 0.732 & 0.717 & 0.708 & 0.712 \\ 
	Iris & 0.748 & 0.757 & 0.69 & 0.7 & 0.772 & 0.760 & 0.763 & 0.757 & 0.761 \\ 
	Gauss\_5\_5\_5 & 0.741 & 0.713 & 0.687 & 0.657 & 0.742 & 0.749 & 0.741 & 0.744 & 0.737  \\ 
	Gauss\_5\_10\_5 & 0.783 & 0.736 & 0.677 & 0.575 & 0.729 & 0.725 & 0.737 & 0.737& 0.729  \\ 
	%Gauss\_5\_50\_5 & 0.646 & 0.596 & 0.621 & 0.570 & 0.622 & 0.619 & 0.627 & 0.631 & 0.625  \\ 
	Gaussian avg & 0.777 & 0.744 & 9.112 & 0.623 & 0.76 & 0.761 & 0.762 & 0.761 & 0.76 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Approx & Approx & Approx & Approx & Approx & Approx & Approx  \\ \hline
	& k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps+ & 3-Swap   \\ \hline
	Soybean & 1.167 & 1.199 & 1.027 & 1.012 & 1.022 & 1.012 & 1.024 \\ 
	Iris & 1.081 & 1.375 & 1.016 & 1.004 & 1.013 & 1.004 & 1.009 \\ 
	Gauss\_5\_5\_5 & 1.070 & 1.436 & 1.030 & 1.002 & 1.027 & 1.005 & 1.026  \\ 
	Gauss\_5\_10\_5 & 1.080 & 1.389 & 1.017 & 1.011 & 1.018 & 1.009 & 1.017  \\ 
	%Gauss\_5\_50\_5 & 1.0887 & 1.272 & 1.030 & 1.020 & 1.027 & 1.014 & 1.033  \\ 
	Gaussian avg & 1.064 & 1.416 & 1.025 & 1.009 & 1.023 & 1.008 & 1.024 \\ 
	pmed5 & 1.373 & 1.681 & 1.06 & 1.026 & 1.071 & 1.033 & 1.051 \\ 
	pmed20 & 1.501 & 1.73 & 1.079 & 1.032 & 1.076 & 1.027 & 1.073 \\ 
	%pmed40 & 1.187 & 1.423 & 1.046 & 1.017 & 1.043 & 1.019 & 1.039 \\ 
	pmedian avg & 1.203 & 1.456 & 1.041 & 1.017 & 1.039 & 1.016 & 1.037 \\ \hline
\end{tabular}

@@

%	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap+ & 2-Swap & 2-swaps+ & 3-Swap & 4-swaps   \\ 

@@

%	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap & 4-swaps   \\ \hline

@@

%	&k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap   \\ \hline

@@

%	  & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps+ & 3-Swap \\ \hline

>>>>

There are two main branches of clustering: partitional (or flat) and hierarchical. In partitional clustering we want to partition the data into disjoint sets whereas in hierarchical clustering we are interested in finding a nested tree of partitions. Often these problems are solved using heuristic algorithms which have no guarantee on the solution produced but are typically fast and easy to implement. We will look at how local search approximation algorithms, which do have some approximation guarantee on the constructed solution and are less computationally expensive than other approximation methods, compare in practice to some of the most widely used clustering algorithms both in the quality of the solution and in computation time.

@@

Arguably the most common partitional clustering method is $k$-means, in which we want to choose $k$ cluster centers to minimize the sum of squared distances from each point to its closest cluster center \cite{Jain}.  $k$-means is usually solved using Lloyd's algorithm (also referred to simply as the $k$-means algorithm), which iteratively greedily updates the centers of each cluster (assigning points to the closest center) until the solution becomes stable \cite{Jain}. Lloyd's algorithm is only guaranteed to reach a locally optimum solution and can be very sensitive to outliers and the initial points chosen \cite{Kanungo}.

@@

Various approximation algorithms for the $k$-median problem have been proposed, with a best known approximation ratio of 2.592 \cite{Wu}. However, this algorithm uses linear programming and is computationally intensive compared to local search and greedy algorithms.  It is known that a local search algorithm that considers swaps of size $\leq p$ is a $(3 + \frac{2}{p})$-approximation algorithm \cite{Arya}. Note that for $p=1$, this is equivalent to PAM.  As far as we know, this algorithm has not been implemented or tested on clustering data, so the relative trade-off of solution quality and computation time as $p$ increases is unknown.

@@

The most common method for hierarchical clustering is agglomerative clustering. Agglomerative clustering starts with each data point in its own cluster, and proceeds to recursively merge the two closest clusters to form the next clustering level in the hierarchy, repeating until the entire data set is a single cluster.  There are several choices on how to measure distance in order to choose which clusters to merge. Single linkage (also called nearest neighbor) defines the distance between two clusters to be the minimum pairwise distance between the data points in each cluster.  Complete linkage clustering instead defines the distance to be the largest pairwise distance. A method between these two extremes is group average linkage which looks at the average pairwise distance between clusters. This strikes a balance between the closeness and compactness of the two techniques above and gives more reasonable results in practice \cite{ESL}. Lastly, there is Ward's method which chooses the two clusters to merge so as to minimize the sum of each cluster's variance \cite{Ward}.

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the flat partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. In particular, their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. The local search $(3+\frac{2}{p})$-approximation algorithm as the $k$-median solver performed well in experimental tests \cite{Nagarajan}. While the authors tested the variants of this algorithm against other $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

@@

\begin{figure*}[h]
\label{fig:flat_results}
\small
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Time & Time & Time & Time & Time & Time & Time & Time & Time \\ \hline
	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap+ & 2-Swap & 2-Swaps+ & 3-Swap  \\ \hline
	Soybean & 0.015 & 0.061 & 0.002 & 1.179 & 0.026 & 0.016 & 0.048 & 0.049 & 0.081 \\ 
	Iris & 0.013 & 0.023 & 0.003 & 9.788 & 0.125 & 0.102 & 0.212 & 0.229 & 0.414 \\ 
	Gauss\_5\_5\_5 & 0.025 & 0.056 & 0.032 & 75.262 & 1.159 & 1.506 & 2.188 & 2.382 & 3.71 \\ 
	Gauss\_5\_10\_5 & 0.054 & 0.095 & 0.109 & 134.137 & 6.834 & 5.407 & 9.243 & 9.428 & 15.96 \\ 
	%Gauss\_5\_50\_5 & 0.137 & 0.196 & 0.302 & 97.336 & 27.856 & 21.446 & 32.407 & 39.614 & 44.69 \\ 
	Gaussian avg & 0.073 & 0.128 & 0.182 & 144.455 & 13.289 & 13.021 & 18.749 & 18.696 & 24.592 \\ 
	pmed5 & 0.038 & 8.298 & 0.481 & 0.622 & 0.653 & 0.748 & 0.939 & \  & \  \\ 
	pmed20 & 1.795 & 174.951 & 52.208 & 44.979 & 68.554 & 68.231 & 79.288 & \  & \  \\ 
	% pmed40 & 3.527 & 849.852 & 251.259 & 264.867 & 250.412 & 391.738 & 350.235 & \  & \  \\ 
	pmedian avg & 0.807 & 203.308 & 40.386 & 40.347 & 48.264 & 56.76 & 56.19 & \  & \  \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI  \\ \hline
	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap+ & 2-Swap & 2-Swaps+ & 3-Swap   \\ \hline
	Soybean & 0.686 & 0.178 & 0.55 & 0.612 & 0.708 & 0.732 & 0.717 & 0.708 & 0.712 \\ 
	Iris & 0.748 & 0.757 & 0.69 & 0.7 & 0.772 & 0.760 & 0.763 & 0.757 & 0.761 \\ 
	Gauss\_5\_5\_5 & 0.741 & 0.713 & 0.687 & 0.657 & 0.742 & 0.749 & 0.741 & 0.744 & 0.737  \\ 
	Gauss\_5\_10\_5 & 0.783 & 0.736 & 0.677 & 0.575 & 0.729 & 0.725 & 0.737 & 0.737& 0.729  \\ 
	%Gauss\_5\_50\_5 & 0.646 & 0.596 & 0.621 & 0.570 & 0.622 & 0.619 & 0.627 & 0.631 & 0.625  \\ 
	Gaussian avg & 0.777 & 0.744 & 9.112 & 0.623 & 0.76 & 0.761 & 0.762 & 0.761 & 0.76 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Approx & Approx & Approx & Approx & Approx & Approx & Approx  \\ \hline
	& k-medoids & Supermodular & PAM & 1-Swap+ & 2-Swap & 2-Swaps+ & 3-Swap   \\ \hline
	Soybean & 1.167 & 1.199 & 1.027 & 1.012 & 1.022 & 1.012 & 1.024 \\ 
	Iris & 1.081 & 1.375 & 1.016 & 1.004 & 1.013 & 1.004 & 1.009 \\ 
	Gauss\_5\_5\_5 & 1.070 & 1.436 & 1.030 & 1.002 & 1.027 & 1.005 & 1.026  \\ 
	Gauss\_5\_10\_5 & 1.080 & 1.389 & 1.017 & 1.011 & 1.018 & 1.009 & 1.017  \\ 
	%Gauss\_5\_50\_5 & 1.0887 & 1.272 & 1.030 & 1.020 & 1.027 & 1.014 & 1.033  \\ 
	Gaussian avg & 1.064 & 1.416 & 1.025 & 1.009 & 1.023 & 1.008 & 1.024 \\ 
	pmed5 & 1.373 & 1.681 & 1.06 & 1.026 & 1.071 & 1.033 & 1.051 \\ 
	pmed20 & 1.501 & 1.73 & 1.079 & 1.032 & 1.076 & 1.027 & 1.073 \\ 
	%pmed40 & 1.187 & 1.423 & 1.046 & 1.017 & 1.043 & 1.019 & 1.039 \\ 
	pmedian avg & 1.203 & 1.456 & 1.041 & 1.017 & 1.039 & 1.016 & 1.037 \\ \hline
\end{tabular}

@@

%	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap+ & 2-Swap & 2-Swaps+ & 3-Swap & 4-swaps   \\ 

@@

%	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-Swaps + & 3-Swap & 4-swaps   \\ \hline

@@

%	&k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-Swaps + & 3-Swap   \\ \hline

@@

%	  & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-Swaps+ & 3-Swap \\ \hline
